Multilevel computation of glacier geometry \\ from Stokes dynamics

### 2021-06-08 paper/msg.tex
Edits sections 2-4.

For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ imply an effective viscosity function of $|D\bu|$.  This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
\begin{equation}
\nu_\eps = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
\end{equation}
where $\pp=(1/\nn)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.

Our SIGP model uses dynamic boundary conditions for isolated, grounded, and non-sliding glaciers.  (Floating and sliding cases are topics for additional research.)  In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.  On the base we impose no slip:
\begin{equation}
\bu = \bzero  \qquad\qquad \text{\emph{base} } \Gamma_0. \label{eq:basebc}
\end{equation}
On the other sub-aerial surfaces we set a condition of zero applied stress,
\begin{equation}
\left(2 \nu_\eps D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{top and cliffs (if present)}} \label{eq:topbc}
\end{equation}
where $\bn$ is any normal to $\overline{\partial} \Lambda$.  (A nonzero atmospheric pressure at the surface is straightforward to apply if desired.)  Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached, but that $\grad s$ may become singular there.

@@

We may now state the strong form of the SIGP, as follows, by including all of the above conditions, and also by eliminating $\tau$ from \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw}:
\begin{align}
s - b &\ge 0 && \text{on $\Omega$} \label{eq:strongform} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{\emph{same}} \notag \\
(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
- \nabla \cdot \left(2 \nu_\eps\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
\nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
\bu &= \bzero && \text{on $\Gamma_0$} \notag \\
\left(2 \nu_\eps D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag
\end{align}
(Definition \eqref{eq:regeffvisc} of $\nu_\eps$ is included.)  The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.

@@

The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here, after which we address the weak form of the SIGP.

Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of the boundary, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/\nn)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=\nn+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $\nn=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
\begin{equation}
\mathcal{M}_\Lambda = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ should satisfy the weak form
\begin{equation}
F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_\Lambda$.

Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.  In particular they show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, but also that there is a unique pressure $p$.

However, we must go beyond the problem on predetermined ice geometry to solve the weak form of \eqref{eq:strongform}.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  From now on we also assume that $\nn>1$.  Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.  (Though it is not clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, the constraint is surely correct.  Existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)

Recalling definition \eqref{eq:lambdas}, we further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
\end{equation}
Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, which we cannot offer.  In the SIA model the surface $s$ solves an elliptic equation, so it is smooth on $\{s>b\} \subset \Omega$ if $b$ is itself smooth \cite{JouvetBueler2012}.

We now define the ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem above (Figure \ref{fig:idoaction}).  For each surface elevation function $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} - \bu|_s \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or have other singularities at the glacier margin, but we will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The ice dynamics operator $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$, equation \eqref{eq:ido}.  Both the input and output of $\Phi$ are defined on all of $\Omega$.}
\label{fig:idoaction}
\end{figure}

We believe that formula \eqref{eq:ido} defines a nonlinear operator $\Phi:\mathcal{K} \subset W^{1,\qq}(\Omega) \to (W^{1,\qq}(\Omega))^* = W^{-1,\pp}(\Omega)$.  This operator is \emph{expensive}; evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}.  Furthermore the operator is clearly \emph{nonlocal} because the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.  In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different everywhere both solutions are defined: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.  Thus an FE discretization of $\Phi(s)$ will have a dense Jacobian matrix

By contrast, in the simpler SIA model the operator is a nonlinear, second-order differential operator,
\begin{equation}
\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right], \label{eq:phisia}
\end{equation}
for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ and $\grad_2 = (\partial_x,\partial_y)$.  Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.  However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid.

We will be concerned with three aspects of the IDO:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item Immediately we use $\Phi$ to simplify the SIGP strong form and express its weak form.
\item Section \ref{sec:smoothers} approximates the diagonal entries of the Jacobian $\Phi'$ to construct smoothers.
\item Section \ref{sec:evolution} relates the spectrum of $\Phi'$ to the stability of time-stepping methods.
\end{enumerate}

For (\emph{i}) we re-write \eqref{eq:strongform} as a simpler NCP with a hidden dynamical problem:
\begin{align}
s - b &\ge 0  \label{eq:idostrongform} \\
\Phi(s) - a &\ge 0 \notag \\
(s - b) (\Phi(s) - a) &= 0 \notag
\end{align}
Here the statements hold on all of $\Omega$ and the roles of the data $a,b$ are evident.  The solution is the ice surface $z=s(x,y)$ while the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.

For $s \in \mathcal{K}$ and $r \in W^{1,\qq}(\Omega)$ we next define a ``residual'' functional which is simply the dual of $\Phi$:
\begin{equation}
F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.

The SIGP weak form is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.  In particular the strong SKE \eqref{eq:ske} holds on $s>b$.

Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a place where $\Phi(s)-a$ points directly into $\mathcal{K}$.  The VI says that the ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.  Also, it is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge VI \eqref{eq:sigpweakform} does \emph{not} arise in this way.

Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.  The general-bed, steady shallow (SIA) model is also not a minimization, except in the flat-bed case \cite{JouvetBueler2012}.

The SIGP weak form \eqref{eq:sigpweakform} is coupled to \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  The VI problem therefore has three fundamental nonlinearities:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item the Glen power-law rheology,
\item the inequality constraint, and
\item the nonlinearity of free-surface flow as a function of the surface.
\end{enumerate}
Regarding (\emph{ii}), observe that the solution of the classical obstacle problem for the linear Laplacian operator is a nonlinear function of the data \cite{KinderlehrerStampacchia1980}.  Regarding (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.

We will need the (Gateaux) derivative of $F$, thus $\Phi$.  Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  From \eqref{eq:sigpfunctional}, if $r\in W^{1,\qq}(\Omega)$ then $F'(s)[t,r] = \int_\Omega \Phi'(s)[t]\, r$ where by definition
\begin{equation}
\Phi'(s)[t] = - \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps}, \label{eq:idoderiv}
\end{equation}
extended by zero to $\Omega$.  However, $\Phi'(s)[t]$ is actually a one-sided directional derivative on a portion of $\Omega$.  That is, $s+\eps t\ge b$ is required for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.  To simplify we define an $s$-independent set
\begin{equation}
\mathcal{W}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, thus $F'(s)[t,r]$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.

The meaning of the difference in \eqref{eq:idoderiv} depends on the support of $t$:
\begin{equation}
\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s = \begin{cases}
           (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b, t > 0 \\
           0, & s=b, t = 0.
                 \end{cases} \label{eq:differencecases}
\end{equation}
Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $s=b$.  In physical terms, the ice motion part of the SKE \eqref{eq:ske} must vanish for small ice masses on bare ground, a reasonable supposition if the ice cannot slide.

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  When we attempt to approximate the derivative $\Phi'(s)[t]$ by a difference quotient, significant performance and stability issues arise.  These are addressed in Section \ref{sec:smoothers}.

>>>>

For the $\nn=1$ (Newtonian) Stokes equations one would write \eqref{eq:viscflowlaw} as $\tau = 2\nu D\bu$ with viscosity $\nu>0$, but exponents $\nn>1$ imply an effective viscosity function of $|D\bu|$.  This would be singular in the limit of small strain rates, and so, motivated by the actual finite viscosity of glacier ice \cite{GreveBlatter2009}, we define the regularized effective viscosity using $\eps>0$,
\begin{equation}
\nu_\eps(|D\bu|) = \frac{1}{2} B_\nn \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}, \label{eq:regeffvisc}
\end{equation}
where $\pp=(1/\nn)+1$, with $\pp=4/3$ in computations.  The constant $D_0$ defines a strain-rate scale for glacier flow; values $D_0 = 1 \,\text{a}^{-1}$ and $\eps = 10^{-4}$ are used in computations.

Our SIGP model uses dynamic boundary conditions for isolated, grounded, and non-sliding glaciers.  (Floating and sliding cases are topics for additional research.)  In addition to the already-stated assumption that the top and bottom boundaries of $\Lambda_s$ can be identified, we further assume these surfaces have well-defined tangents.  On the base we impose no slip:
\begin{equation}
\bu = \bzero  \qquad\qquad \text{\emph{base} } \Gamma_0. \label{eq:basebc}
\end{equation}
On the other sub-aerial surfaces we set a condition of zero applied stress,
\begin{equation}
\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn = \bzero  \qquad \qquad \text{\emph{top and cliffs (if present)}} \label{eq:topbc}
\end{equation}
where $\bn$ is any normal to $\overline{\partial} \Lambda$.  (A nonzero atmospheric pressure at the surface is straightforward to apply if desired.)  Note that the ice flow extends in the horizontal direction until a free boundary at the glacier margin is reached, but that $\grad s$ may become singular there.

@@

We may now state the strong form of the SIGP, as follows, by including all of the above conditions, and also by eliminating $\tau$ from \eqref{eq:forcebalance} and \eqref{eq:viscflowlaw}:
\begin{align}
s - b &\ge 0 && \text{on $\Omega$} \label{eq:strongform} \\
- \bu|_s \cdot \bn_s - a &\ge 0 && \text{\emph{same}} \notag \\
(s - b) (- \bu|_s \cdot \bn_s - a) &= 0 && \text{\emph{same}} \notag \\
- \nabla \cdot \left(2 \nu_\eps(|D\bu|)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{on $\Lambda_s$} \notag \\
\nabla \cdot \bu &= 0 && \text{\emph{same}} \notag \\
\bu &= \bzero && \text{on $\Gamma_0$} \notag \\
\left(2 \nu_\eps(|D\bu|) D\bu - pI\right) \bn &= \bzero && \text{on $\partial \Lambda_s \setminus \Gamma_0$} \notag
\end{align}
The first three statements in \eqref{eq:strongform} form the NCP, but this problem is coupled to the boundary value problem formed by the last four statements.

@@

The weak form of the fixed-geometry Glen-Stokes model for ice flow, which extends the linear Stokes weak form \cite{Elmanetal2014} to regularized power-law rheology, is relatively well-known \cite{IsaacStadlerGhattas2015,JouvetRappaz2011,Lengetal2012}, and we summarize it here.  Then we address the weak form of the SIGP.

Denote by $W^{k,r}$ the Sobolev space \cite{Evans2010} of functions with $k$ weak derivatives which are $r$th-power integrable.  Suppose for the moment that $\Lambda \subset \RR^{d+1}$ is a fixed, bounded domain on which we prescribe a Dirchlet condition $\bu=0$ on some positive-measure portion of $\partial\Lambda$, and otherwise suppose the boundary is stress-free (Neumann boundary).  Let $\pp=(1/\nn)+1$ as in \eqref{eq:regeffvisc}, and $\qq=(1-\pp^{-1})^{-1}=\nn+1$ the conjugate exponent; $\pp=4/3$ and $\qq = 4$ if $\nn=3$.  Let $W_0^{1,\pp}(\Lambda)^{d+1}$ be the space of velocity functions, zero along the Dirichlet boundary, and let
\begin{equation}
\mathcal{M}_\Lambda = W_0^{1,\pp}(\Lambda)^{d+1} \times L^\qq(\Lambda)  \label{eq:mixed}
\end{equation}
be the (mixed) space of admissible velocity and pressure pairs.  The Glen-Stokes solution $(\bu,p) \in \mathcal{M}_\Lambda$ satisfies the weak form
\begin{equation}
F_\Lambda(\bu,p)[\bv,q] = \int_\Lambda 2 \nu_\eps(|D\bu|) D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0 \label{eq:glenstokesweak}
\end{equation}
for all $(\bv,q) \in \mathcal{M}_\Lambda$.  If $\bu,p$ are sufficiently regular then they satisfy strong equations \eqref{eq:forcebalance}--\eqref{eq:topbc}.

Jouvet and Rappaz \cite{JouvetRappaz2011} prove that this fixed-domain Glen-Stokes formulation is well-posed under the above assumptions if also the Neumann boundary of $\Lambda$ is $C^1$.  They show \eqref{eq:glenstokesweak} is equivalent to minimization of a convex and coercive functional over the divergence-free subspace, and that there is a unique pressure $p$.

However, we must go beyond the problem on predetermined ice geometry to solve the weak form of SIGP \eqref{eq:strongform}.  Again suppose $\Omega \subset \RR^d$ is a fixed map-plane domain with $C^1$ boundary, and that the bed elevation $b$ is in $W^{1,\qq}(\Omega)$.  The Sobolev space $W^{1,\qq}(\Omega)$ supports a linear trace operator which defines the symbol $|_{\partial \Omega}$ \cite[Section 5.5]{Evans2010}.  Let
\begin{equation}
\mathcal{K} = \{s \in W^{1,\qq}(\Omega) \,:\, s \ge b \, \text{ and } \, s\big|_{\partial\Omega} = b\big|_{\partial\Omega}\}  \label{eq:Kconstraintset}
\end{equation}
be the closed and convex set of admissible surface elevations of an isolated glacier or ice sheet.  From now on we also assume that $\nn>1$.  Because $\qq = \nn+1 > d$, each $s \in W^{1,\qq}(\Omega)$ is continuous \cite[Morrey's inequality, section 5.6.2]{Evans2010}, and so there are no cliffs in the $s$ geometry.  (Though it is not yet clear that the correct Sobolev space is identified in \eqref{eq:Kconstraintset}, existence for the simpler SIA model finds a power of the ice thickness, $u=(s-b)^{2q/(q-1)}$, in $\mathcal{K} = \{v \ge 0\} \subset W^{1,\qq}(\Omega)$ \cite{JouvetBueler2012}.)  Recalling definition \eqref{eq:lambdas}, we further suppose that
\begin{equation}
s\in \mathcal{K} \text{ defines an icy domain } \Lambda_s \text{ which has a $C^1$ upper surface.} \label{eq:quixotic}
\end{equation}
Assumption \eqref{eq:quixotic} would follow from a sufficiently-strong regularity result for the theory of the current paper, something we cannot offer.  However, in the SIA model the surface $s$ solves an elliptic equation, and $s$ is $C^1$ on $\{s>b\} \subset \Omega$ if $b$ is continuous \cite{JouvetBueler2012}.

We now define an ice dynamics operator (IDO), a map $\Phi$ which uses the trace of the velocity solution to the Glen-Stokes problem \eqref{eq:glenstokesweak}; see Figure \ref{fig:idoaction}.  For each $s \in \mathcal{K}$ the output $\Phi(s)$ is the negative normal component of the (unique) velocity solution $\bu$ of \eqref{eq:glenstokesweak} over the upper surface of $\Lambda_s$, but extended by zero to all of $\Omega$:
\begin{equation}
\Phi(s) = \begin{cases} - \bu|_s \cdot \bn_s, & s > b, \\
                        0, & \text{otherwise}. \end{cases} \label{eq:ido}
\end{equation}
Here $\bu|_s$ denotes a trace operator on $W^{1,\pp}(\Lambda_s)^{d+1}$; compare \eqref{eq:mixed}.  Note that $\Phi(s)$ is a scalar function on $\Omega$ which may jump or be singular at the glacier margin even if $s$ is smooth on $\{s>b\}$.  (We will only need $\Phi(s)$ to be in the dual of $W^{1,\qq}(\Omega)$.)

\begin{figure}[t]
\begin{center}
\includegraphics[width=\textwidth]{genfigs/idoaction.pdf}
\end{center}
\caption{The IDO $\Phi$ maps the surface elevation $s$ (dashed) to the normal ice surface motion $\Phi(s)=- \bu|_s \cdot \bn_s$.  Both $s$ and $\Phi(s)$ are defined on all of $\Omega$, but $\Phi(s)$ may be discontinuous as shown.}
\label{fig:idoaction}
\end{figure}

We believe that formula \eqref{eq:ido} defines a nonlinear operator $\Phi:\mathcal{K} \subset W^{1,\qq}(\Omega) \to (W^{1,\qq}(\Omega))^* = W^{-1,\pp}(\Omega)$.  This operator is \emph{expensive} and \emph{nonlocal}.  Evaluating $\Phi(s)$ requires solving Glen-Stokes problem \eqref{eq:glenstokesweak} on $\Lambda_s \subset \RR^{d+1}$ defined in \eqref{eq:lambdas}, and furthermore the solution $\bu$ is influenced everywhere by changes in the icy domain geometry.  In particular, if $\tilde s=s + r$ defines $\Lambda_{\tilde s} \supset \Lambda_s$, for $r\in C_c^\infty(\Omega)$ with $r\ge 0$ supported inside $\{s>b\}$, then the Stokes solutions are generically different \emph{everywhere}: $\tilde\bu - \bu \ne 0$ on all of $\Lambda_s$, including the trace on the upper surface.  It follows that an FE discretization (Section \ref{sec:fe}) of $\Phi(s)$ will have a dense Jacobian matrix

By contrast, in the simpler isothermal, nonsliding SIA model the operator is a nonlinear, second-order differential operator,
\begin{equation}
\Phi_{\text{SIA}}(s) = - \frac{\gamma}{\qq} (s-b)^{\qq} |\grad_2 s|^{\qq} - \grad_2 \cdot\left[\frac{\gamma}{\qq+1} (s-b)^{\qq+1} |\grad_2 s|^{\qq-2} \grad_2 s\right], \label{eq:phisia}
\end{equation}
for $\gamma = 2(\rhoi g)^{\nn} (B_\nn)^{-\nn} > 0$ constant and $\grad_2 = (\partial_x,\partial_y)$.  Thus an FE discretization of $\Phi_{\text{SIA}}(s)$ has sparse Jacobian.  However, $\Phi_{\text{SIA}} \approx \Phi$ for shallow glaciers to the degree that the lubrication approximation \cite{Acheson1990} is valid, and in this sense we expect the Jacobian of $\Phi$ to be well-approximated by a banded matrix.

In this section we will use $\Phi$ to simplify the SIGP strong form and express its weak form, but later we will be concerned with two other aspects of the IDO:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item Section \ref{sec:smoothers} approximates the diagonal entries of the Jacobian $\Phi'$ to construct smoothers.
\item Section \ref{sec:evolution} relates the spectrum of $\Phi'$ to the stability of time-stepping methods.
\end{enumerate}

For now we re-write \eqref{eq:strongform} as a clean NCP with a hidden dynamical problem:
\begin{align}
s - b &\ge 0  \label{eq:idostrongform} \\
\Phi(s) - a &\ge 0 \notag \\
(s - b) (\Phi(s) - a) &= 0 \notag
\end{align}
Here the statements hold on all of $\Omega$ and the roles of the data $a,b$ are evident.  The solution is the ice surface $z=s(x,y)$ while the variables $\bu,p$ are hidden within the evaluation of $\Phi(s)$.

To show the SIGP weak form we define a functional which is simply the dual of $\Phi$:
\begin{equation}
F(s)[r] = \ip{\Phi(s)}{r} = \int_\Omega \Phi(s)\, r \,dx dy. \label{eq:sigpfunctional}
\end{equation}
One computes $F(s)[r]$ by defining $\Lambda_s$ as in \eqref{eq:lambdas}, solving \eqref{eq:glenstokesweak}, evaluating the surface trace of the velocity to compute $\Phi(s)$, and then integrating the result against (i.e.~dual pairing with) the test function $r$.  The weak form itself is the following variational inequality (VI) \cite{KinderlehrerStampacchia1980} for $s\in\mathcal{K}$:
\begin{equation}
F(s)[r - s] \ge \ip{a}{r-s} \quad \text{for all $r \in \mathcal{K}$.}  \label{eq:sigpweakform}
\end{equation}
One can prove that if $s$ is $C^1$ and solves \eqref{eq:sigpweakform} then it solves strong forms \eqref{eq:strongform} and \eqref{eq:idostrongform} \cite{Bueler2021conservation}.  In particular the strong SKE \eqref{eq:ske} holds on $s>b$; it is the interior condition of VI \eqref{eq:sigpweakform} \cite{KinderlehrerStampacchia1980}.

Speaking geometrically, \eqref{eq:sigpweakform} says that $s$ is located in $\mathcal{K}$, generically on $\partial\mathcal{K}$, at a point where $\Phi(s)-a$ points directly into $\mathcal{K}$.  (The ``angle'' between $\Phi(s)-a$ and an arbitrary vector $r-s$ pointing into $\mathcal{K}$ is at most $90^\circ$.)  It is well-known that certain VIs arise as inequality-constrained minimization problems \cite{GraeserKornhuber2009,KinderlehrerStampacchia1980}, but to the best of our knowledge \eqref{eq:sigpweakform} does \emph{not} arise in this way.  Existence has been proven for the SIA analog of VI \eqref{eq:sigpweakform} \cite{JouvetBueler2012}, but, as noted above, in that case the residual is locally-computable.  This model is also not a minimization, except in the flat-bed case.

The SIGP weak form \eqref{eq:sigpweakform} is coupled to the Glen-Stokes weak form \eqref{eq:glenstokesweak} through definition \eqref{eq:ido} of the IDO $\Phi$.  VI problem \eqref{eq:sigpweakform} therefore has three fundamental nonlinearities:
\renewcommand{\labelenumi}{(\emph{\roman{enumi}})}
\begin{enumerate}
\item the Glen power-law rheology,
\item the inequality constraint, and
\item the nonlinearity of free-surface flow as a function of the surface.
\end{enumerate}
Regarding (\emph{ii}), observe that the solution of the classical obstacle problem for the linear Laplacian operator is a nonlinear function of the data \cite{KinderlehrerStampacchia1980}.  For (\emph{iii}), note that a free-surface flow for Newtonian (linear) rheology produces a nonlinear equation for the flow thickness, e.g.~as expressed in the kinematic wave equation \cite{Ockendonetal2003}.

We will need the (Gateaux) derivative of $\Phi$.  Suppose $s\in \mathcal{K}$, $\eps>0$, and $t \in W_0^{1,\qq}(\Omega)$ is such that $s+\eps t \in \mathcal{K}$.  From \eqref{eq:sigpfunctional}, if $r\in W^{1,\qq}(\Omega)$ then by definition
\begin{equation}
\Phi'(s)[t] = - \lim_{\eps\to 0^+} \frac{\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s}{\eps} \label{eq:idoderiv}
\end{equation}
on $\{s>b\}$.  However, $\Phi'(s)[t]$ must actually be defined as a one-sided directional derivative on $\{s=b\}$.  That is, $s+\eps t\ge b$ is required for all sufficiently-small $\eps>0$, which requires $t\ge 0$ on the (active) set where $s=b$.  To simplify we define an $s$-independent set
\begin{equation}
\mathcal{W}_+ = \{t \in W_0^{1,\qq}(\Omega) \,:\, t(x,y) \ge 0\}. \label{eq:infdefectset}
\end{equation}
It follows that $\Phi'(s)[t]$ is well-defined on inputs $(s,t) \in \mathcal{K} \times \mathcal{W}_+$, thus $F'(s)[t,r]=\int_\Omega \Phi'(s)[t] r$ is well-defined for $(s,t,r) \in \mathcal{K} \times \mathcal{W}_+ \times W^{1,\qq}(\Omega)$.

The difference in \eqref{eq:idoderiv} depends on the support of $t$:
\begin{equation}
\bu|_{s+\eps t} \cdot \bn_{s+\eps t} - \bus \cdot \bn_s = \begin{cases}
           (\bu|_{s+\eps t} - \bus) \cdot \bn_{s+\eps t} + \eps\, \bus \cdot \left<t_x,t_y,0\right>, & s > b \\
           \bu|_{s+\eps t} \cdot \bn_{s+\eps t}, & s=b, t > 0 \\
           0, & s=b, t = 0.
                 \end{cases} \label{eq:differencecases}
\end{equation}
Thus a necessary condition for differentiability of $\Phi$ is that $\bu|_{s+\eps t} \cdot \bn_{s+\eps t} = o(\eps)$ on the ice-free area $\{s=b\}$.  In physical terms, the ice motion part of the SKE \eqref{eq:ske} must vanish for shrinking ice masses on bare ground, a reasonable supposition if the ice cannot slide.

In the next three sections we will construct an iterative, multilevel finite element solver for SIGP weak form \eqref{eq:sigpweakform}.  Substantial performance and stability concerns related to evaluating $\Phi'$ are addressed in Section \ref{sec:smoothers}.
